Decoupled Fallback: A Portable Single-Pass GPU Scan

### 2025-02-12 paper.tex
Add Apple data, rescale graphics.

\section{Background}
\subsection{Virtualization and Scheduling in the GPU programming model}
Contemporary GPUs are hierarchically organized, massively parallel processors designed to prioritize throughput over latency. (one sentence on thread hierarchy maybe). As comprehensive descriptions of the GPU programming model~\cite{} already exist, we focus on the aspects most relevant to this work: scheduling and synchronization.

\raph{I think "NVIDIA Tesla: A Unified Graphics and Computing Architecture" Lindholm 2008 might be a good choice for the above cite.}
\john{I have historically used that Lindholm reference for the GPU hardware, but for the programming model I have used Nickolls, ``Scalable parallel programming with CUDA'' (ACM Queue, Mar/Apr 2008).}

@@

\subsubsection{Chained-Scan, Stream-Scan}
First introduced by Yan et al.~\cite{10.1145/2442516.2442539} in \emph{Stream-Scan}, the key innovation of the \emph{Chained-Scan} architecture lies in its hybridization of parallel and serial strategies: achieving parallelism at the intra-workgroup level while minimizing global data movement through serial scan operations at the inter-workgroup level. Instead of using kernel launches as inter-workgroup synchronization points, \emph{Chained-Scans} launch a single kernel that assigns work tiles in a serial order to workgroups, utilizing atomics and bit-packed status flags to ensure coherent views of data. To enforce this serial ordering, work tiles are dynamically assigned using atomic increment operations, rather than arbitrary assignment based on virtualized workgroup indices. This shifts dependency resolution from explicit synchronization via barrier-like structures to implicit synchronization via access order. As a result, each element is read and written exactly once, achieving the theoretical minimum of $2n$.

@@

For example, consider a scan operation requiring two workgroups on a GPU with a single multiprocessor. Both workgroups are co-resident, but the multiprocessor has only one scheduling unit, which lacks FPG\@. Both workgroups launch and acquire their respective work tiles. The workgroup responsible for \emph{Work Tile 1} finishes first and begins spinning, repeatedly querying the memory address for the reduction of \emph{Work Tile 0}. However, because the scheduler lacks FPG, it continues scheduling \emph{Work Tile 1}, leaving \emph{Work Tile 0} starved and unable to complete its reduction. As a result, \emph{Work Tile 1} spins indefinitely, eventually triggering \emph{timeout detection and recovery}\footnote{\emph{Timeout detection and recovery} is} (TDR) and crashing the host program.

@@

In particular, we use two or four 32-bit values, splitting the reduction value into 16-bit segments while reserving the two most significant bits of each 32-bit value to encode the tile state.
\begin{figure}[h!]
  \centering
  \includegraphics[width=\linewidth]{graphics/split.pdf}
\end{figure}

To accommodate the split values during \emph{Chained Propagation}, we use two or four threads from the first subgroup, leveraging subgroup \emph{ballot} to verify state consistency, and subgroup \emph{shuffle} to join and split reductions. With one thread per 16 bits and a minimum subgroup size of four in WebGPU, this method supports both 32-bit and 64-bit values, enabling 64-bit scans without 64-bit atomics or fences.\footnote{Although native 64-bit values are not currently supported by the WebGPU specification, their future inclusion is anticipated.}

@@

\section{Evaluation}
\subsection{Experimental Setup}
We configure workgroups with a size $w = 256$ threads, where each thread processes $p = 4$ vectors of size 4, resulting in 16 elements per thread and a total work tile size of $t = 4096$. Through testing, \john{across all devices (?)} \thomas{Only rigorously tested on the 2080, everything else was very light.}we find a maximum spin count of 4 effectively balances responsiveness to blocking while minimizing false-positive fallbacks. As the size of the local spine is inversely proportional to the subgroup size, we allocate sufficient shared memory to accommodate the WebGPU standard's minimum subgroup size of 4. Due to our overall thriftiness with shared memory, total usage remains modest at only $\sim$1 KB\@. For all tests, we use an input size of $2^{25}$ and an inclusive prefix sum. Our test hardware, listed POSITIONAL-RELATION, spans a wide range of FPG and performance capabilities. We empirically find that the Google Chrome native implementation of WebGPU (Dawn) provides the best performance and use it for all testing, except on the Arm-Mali GPU, which requires Android native, necessitating the use of WGPU\@.\footnote{To keep the perforamance as close to Dawn as possible, we translate our WGSL code with Dawn's Tint transpiler, and provide the raw SPIR-V to WGPU\@.}
\begin{table}
  \centering
  \small
  \begin{tabular*}{\linewidth}{@{\extracolsep{\fill}} l l c r r}
    \toprule
    Vendor & Device Name & OBE? & Occupancy & Subgroup Size \\
    \midrule
    Intel  & HD Graphics 620       & \textbf{Yes} & 9   & 8--32  \\
    ARM    & Mali-G78 MP20          & \textbf{Yes} & 40  & 16  \\
    Apple  & M1 Pro Max             & \textbf{No}  & 64  & 32  \\
    NVIDIA & GeForce RTX 2080 Super & \textbf{Yes} & 192 & 32  \\
    AMD    & Radeon RX 7900 XT      & \textbf{Yes} & 420 & 32--64  \\
    \bottomrule
  \end{tabular*}
  \caption{Hardware Specifications of Testing Devices.}
  \label{tab:hardware}
\end{table}

\subsection{Performance Comparison}
\subsubsection{Performance Versus Contemporary Implementations:}
We first compare \emph{Decoupled Fallback} (\emph{CSDLDF}) to the \emph{Decoupled Lookback} (\emph{CSDL}) implementation found in NVIDIA’s \emph{CUB} library, using the 2080 Super only. To assess the impact of differing APIs, we also implement \emph{CSDLDF} in CUDA.\footnote{Due to WebGPU’s 128MB storage buffer limit, our maximum test size is limited to $2^{25}$ 32-bit elements.}

Cross-API comparisons remain challenging, as despite being nearly identical at a high level, the CUDA implementation consistently outperforms its WebGPU counterpart. Additionally, the performance cost of \emph{Decoupled Fallback} is most evident in our CUDA implementation when the entire input fits within the 2080 Super’s 4MB L2 cache, where increased synchronization and compute overhead slightly reduce throughput. However, at larger input sizes, where memory accesses extend beyond the cache, \emph{CSDLDF} achieves parity with \emph{CUB}.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{graphics/cuda_plot.pdf}
  \caption{Performance versus \emph{CUB} (NVIDIA 2080 Super)}
\end{figure}

@@

\begin{table*}
  \centering
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l r r r r r}
    \toprule
    Metric & Intel HD620 & ARM Mali-G78 MP20 & Apple M1 Pro Max & Nvidia 2080 Super & AMD 7900 XT \\
    \midrule
    H-Mean \emph{Memcpy} (32-bit ele/sec)  & 1.498e9   & 106.663e9 & 54.203e9  & 54.203e9  & 71.190e9  \\
    H-Mean \emph{CSDLDF} (32-bit ele/sec)   & 1.487e9   & 112.430e9 & 53.256e9  & 53.256e9  & 95.737e9  \\
    H-Mean \emph{RTS} (32-bit ele/sec)      & 1.060e9   & 82.302e9  & 35.709e9  & 35.709e9  & 62.868e9  \\
    \emph{CSDLDF}/\emph{Memcpy} (\%)       & 99.2\%    & 105.4\%   & 98.3\%    & 98.3\%    & 134.5\%   \\
    \emph{CSDLDF}/\emph{RTS} (\%)          & 140.2\%   & 136.6\%   & 149.1\%   & 149.1\%   & 152.3\%   \\
    Avg Spins per Wg    & 0.306    & 1.380    & 0.418    & 0.418    & 0.133    \\
    Avg Lookback Length per Wg  & 1.725    & 3.510    & 13.139   & 13.139   & 20.867   \\
    Avg Total Fallbacks Initiated      & 260.364   & 2202.045  & 88.384   & 88.384   & 10.127   \\
    Avg Total Successful Insertions    & 6.084     & 57.875    & 1.205    & 1.205    & 1.001    \\
    \bottomrule
  \end{tabular*}
  \caption{MALI ABSOLUTE SPEED WRONG!, \emph{Decoupled Fallback} performance, 32-bit Inclusive Prefix Sum, Input Size $\mathbf{2^{25}}$. \john{I would label metrics as ``H-Mean \emph{Memcpy} (32-bit ele/sec $\cdot 10^9$)'' and then leave the exponents out of the main table.}\john{Separately, h-mean is the right summary statistic \emph{if} the distribution is Gaussian, which it almost certainly is not. Like most measurements of this type, it's much more likely log-normal. I am influenced here by \url{https://lemire.me/blog/2023/04/06/are-your-memory-bound-benchmarking-timings-normally-distributed/}.}}
  \label{tab:results}
\end{table*}

As shown in Table~\ref{tab:results}, with the exception of the ARM Mali, \emph{CSDLDF} achieves at least a $\sim$40\% speedup over \emph{RTS} across all tested devices. Since \emph{Memcpy} is purely memory bandwidth bound, \emph{CSDLDF} should ideally match its throughput. This holds across most devices, except for the AMD 7900 XT, where \emph{CSDLDF} outperforms \emph{Memcpy}, suggesting that the additional compute latency counterintuitively improves memory access patterns by reducing paging. \john{I have no idea what this previous phrase means.} \thomas{In my head, if you read from location A, and then write immediately to B, you need room in your cache for A and B. If you read, do just the right amount of work, then write B, you only need to hold A or B? Not sure if this is correct.} On the ARM Mali, \emph{CSDLDF} appears to underperform, achieving only a 36\% speedup instead of the expected 50\%. However, as its performance is nearly identical to \emph{Memcpy}, this indicates that the implementation is memory-bandwidth-bound, limiting further gains.

\subsection{Decoupled Fallback Statistics}
To better understand how devices with differing schedulers interact with \emph{CSDLDF}, we also collect a number of statistics during its execution:
\begin{itemize}
  \item \textbf{Spins per Workgroup:}
  \item \textbf{Lookback Length per Workgroup:}
  \item \textbf{Total Fallbacks Initiated:}
  \item \textbf{Total Successful Insertions:}
\end{itemize}

\subsection{Simulated Blocking}
To demonstrate the resilience of our technique under fallback rates significantly higher than those observed organically on devices, we intentionally force fallbacks by omitting the posting of work tile reductions. Starting with infrequent omissions---1 in 512 workgroups---we progressively increase the rate to near-continuous levels, reaching 1 in 2 workgroups.
\begin{figure*}[t]
  \centering
  \begin{subfigure}[t]{0.48\linewidth}
    \centering
    \includegraphics[width=\linewidth]{graphics/fallbacksInitiated_plot.pdf}
    \caption{Fallbacks Initiated}
    \label{fig:fallbacks_initiated}
  \end{subfigure}\hfill
  \begin{subfigure}[t]{0.48\linewidth}
    \centering
    \includegraphics[width=\linewidth]{graphics/totalSpins_plot.pdf}
    \caption{Total Spins}
    \label{fig:total_spins}
  \end{subfigure}

@@

In Figure~\ref{fig:fallbacks_initiated}, we measure the number of fallbacks proportional to the number of workgroups launched, and verify that we are indeed inducing fallbacks. In Figure~\ref{fig:total_spins} we measure \thomas{this should be spins per workgroup it's a better metric}. In Figure~\ref{fig:lookback_length} we see that. Finally, in Figure~\ref{fig:execution_time}, we can see that, even when \emph{half} the workgroups are deadlocking, \emph{Decoupled-Fallback} still achieves speeds of . . .

\section{Discussion}

\subsection{Trade-offs}

\subsection{Limits on Speed-of-Light Performance}
There are a number of factors which may preclude speed-of-light performance, but most salient is the fairness of the scheduling model. In \emph{Decoupled-Fallback}, a work tile which is late due to unfairness is indistinguishable from one which is blocking, so as a scheduler becomes increasingly unfair, it incurs an increasing number of redundant fallbacks. Consider a hardware with a workgroup occupancy denoted by $o$, and let $f$ represent the probability that a fallback operation occurs at a particular lookback step. Because the number of lookback steps is bounded by $o$, the expected number of fallback operations is approximately $fo$. This results in a factor of $fo$ increase in global memory reads and a factor of $fo\log{n}$ increase in work.

This increased sensitivity to fairness exacerbates existing issues which negatively impact the performance of previous scan architectures, namely lack of compute, and slow atomic update latency. On hardware that lacks sufficient compute power to be memory-bandwidth bound, the additional work incurred by redundant fallback reductions is particularly deleterious. High atomic update latency further compounds the problem: as inter-workgroup communication time grows, the minimum delay before updates become visible to dependent workgroups also increases. This, in turn, raises the number of lookback steps that may be needed and potentially leads to redundant fallbacks.

@@

@@

\clearpage
\appendix
\section{Artifact}
\subsection{Baseline Comparisons: \emph{Memcpy}}
\label{sec:memcpy}
\begin{table}[h!]
  \small
  \centering
  \begin{tabular}{l r r}
    \toprule
    Device                                & \makecell{Memory                   \\ Bandwidth \\ (GB/s)} & \makecell{Expected \\ Throughput \\ (32-bit ele/sec)} \\
    \midrule
    Intel HD620 (Single-/Dual-Channel)    & 17.1 / 34.1      & 2.13e9 / 4.26e9 \\
    ARM Mali-G78 MP20 / Pixel 6a (Tensor) & 51.2             & 6.40e9          \\
    Apple M1 Pro Max                      & 200              & 25.0e9          \\
    Nvidia 2080 Super                     & 496              & 62.0e9          \\
    AMD 7900 XT                           & 800              & 100.0e9         \\
    \bottomrule
  \end{tabular}
  \caption{Vendor-Specified Memory Bandwidth and Expected Throughput (32-bit Elements per Second)}
  \label{tab:memory_bandwidth}
\end{table}
Although all backend APIs support \emph{timestamp queries inside encoders}, Apple GPUs do not, so we are forced to implement our own \emph{Memcpy} kernel to measure memory bandwidth, as opposed to simply timing an API's copy buffer operation. It is well known that \emph{observed} peak memory bandwidth, tends to diverge from vendor-listed peak bandwidth, so for completeness we list these figures above. We make a number of observations:
\begin{itemize}
  \item \textbf{Intel HD620:} As our HD620 laptop is only equipped with a single memory DIMM, its peak bandwidth is limited to the lower bound, 2.13e9 ele/sec. Thus, our observed bandwidth of 1.498e9 is reasonable.
  \item \textbf{ARM Mali-G78 MP20 / Pixel 6a (Tensor):} The observed throughput is significantly lower than expected, despite a vendor-listed bandwidth of 6.40e9 ele/sec. The cause of this discrepancy remains unclear, but we note other developers have observed similar speeds.
  \item \textbf{Apple M1 Pro Max:} Observed throughput matches expectations.
  \item \textbf{Nvidia 2080 Super:} Observed throughput matches expectations.
  \item \textbf{AMD 7900 XT:} Our above result, where \emph{CSDLDF} was observed to be faster than \emph{Memcopy}, is likely due to compute latency counterintuitively improving memory access patterns, leading to reduced paging. Furthermore, we see that our \emph{CSDLDF} throughput is within spec and reasonable.
\end{itemize}

>>>>

\section{Background}
\subsection{Virtualization and Scheduling in the GPU programming model}
Contemporary GPUs are hierarchically organized, massively parallel processors designed to prioritize throughput over latency. As comprehensive descriptions of the GPU programming model~\cite{10.1145/1365490.1365500} already exist, we focus on the aspects most relevant to this work: scheduling and synchronization.

@@

\subsubsection{Chained-Scan, Stream-Scan}
First introduced by Yan et al.~\cite{10.1145/2442516.2442539} in \emph{Stream-Scan}, the key innovation of the \emph{Chained-Scan} architecture lies in its hybridization of parallel and serial strategies: achieving parallelism at the intra-workgroup level while minimizing global data movement through serial scan operations at the inter-workgroup level. Instead of using kernel launches as inter-workgroup synchronization points, \emph{Chained-Scans} launch a single kernel that assigns work tiles in a serial order to workgroups, utilizing atomics and bit-packed status flags to ensure coherent views of data. To enforce this serial ordering, work tiles are dynamically assigned using atomic increment operations, rather than arbitrary assignment based on virtualized workgroup indices. This shifts dependency resolution from explicit synchronization via barrier-like structures to implicit synchronization via access order. As a result, each element is read and written exactly once, achieving the theoretical minimum of $O(2n)$.

@@

For example, consider a scan operation requiring two workgroups on a GPU with a single multiprocessor. Both workgroups are co-resident, but the multiprocessor has only one scheduling unit, which lacks FPG\@. Both workgroups launch and acquire their respective work tiles. The workgroup responsible for \emph{Work Tile 1} finishes first and begins spinning, repeatedly querying the memory address for the reduction of \emph{Work Tile 0}. However, because the scheduler lacks FPG, it continues scheduling \emph{Work Tile 1}, leaving \emph{Work Tile 0} starved and unable to complete its reduction. As a result, \emph{Work Tile 1} spins indefinitely, potentially triggering \emph{timeout detection and recovery}\footnote{Operating systems have different mechanisms for handling unresponsive programs. We borrow here Window's terminology. On Windows, the Timeout Detection and Recovery system resets the GPU after 2 seconds by default. iOS has a timeout watchdog, which, interestingly, does not terminate a spinning kernel. On Linux, the behavior depends on the GPU driver.} (TDR) and crashing the host program.

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=\linewidth]{graphics/split.pdf}
\end{figure}

In particular, we use two or four 32-bit values, splitting the reduction value into 16-bit segments while reserving the two most significant bits of each 32-bit value to encode the tile state. Although this could also be accomplished with shared memory (and barriers), as an optimization, we use two or four threads from the first subgroup, leveraging subgroup \emph{ballot} to verify state consistency, and subgroup \emph{shuffle} to join and split reductions. With one thread per 16 bits and a minimum subgroup size of four in WebGPU, this method supports both 32-bit and 64-bit values, enabling 64-bit scans without 64-bit atomics or fences.\footnote{Although native 64-bit values are not currently supported by the WebGPU specification, their future inclusion is anticipated.}

@@

\section{Evaluation}
\subsection{Experimental Setup}
We configure workgroups with a size $w = 256$ threads, where each thread processes $p = 4$ vectors of size 4, resulting in 16 elements per thread and a total work tile size of $t = 4096$. Through testing, \john{across all devices (?)} \thomas{Only rigorously tested on the 2080, everything else was very light.}we find a maximum spin count of 4 effectively balances responsiveness to blocking while minimizing false-positive fallbacks. As the size of the local spine is inversely proportional to the subgroup size, we allocate sufficient shared memory to accommodate the WebGPU standard's minimum subgroup size of 4. Due to our overall thriftiness with shared memory, total usage remains modest at only $\sim$1 KB\@. For all tests, we use an input size of $2^{25}$ and an inclusive prefix sum. Our test hardware, listed POSITIONAL-RELATION, spans a wide range of FPG and performance capabilities. We empirically find that the Google Chrome native implementation of WebGPU (Dawn) provides the best performance and use it for all testing, except on the Arm-Mali GPU, which requires Android native, necessitating the use of WGPU\@.\footnote{To keep the perforamance as close to Dawn as possible, we translate our WGSL code with Dawn's Tint transpiler, and provide the raw SPIR-V to WGPU\@.}
\begin{table}
  \centering
  \small
  \begin{tabular*}{\linewidth}{@{\extracolsep{\fill}} l l c r r}
    \toprule
    Vendor & Device Name & OBE & Occupancy & Subgroup Size \\
    \midrule
    Intel  & HD Graphics 620        & \textbf{Yes} & 9   & 8--32  \\
    ARM    & Mali-G78 MP20          & \textbf{Yes} & 40  & 16  \\
    Apple  & M3                     & \textbf{No}  & 80  & 32  \\
    Apple  & M1 Max                 & \textbf{No}  & 160 & 32  \\
    NVIDIA & GeForce RTX 2080 Super & \textbf{Yes} & 192 & 32  \\
    AMD    & Radeon RX 7900 XT      & \textbf{Yes} & 420 & 32--64  \\
    \bottomrule
  \end{tabular*}
  \caption{Hardware Specifications of Testing Devices.}
  \label{tab:hardware}
\end{table}

\subsection{Performance Comparison}
\subsubsection{Performance Versus Contemporary Implementations:}
We first compare \emph{Decoupled Fallback} (\emph{CSDLDF}) to the \emph{Decoupled Lookback} (\emph{CSDL}) implementation found in NVIDIA’s CUB library, using the 2080 Super only. To assess the impact of differing APIs, we also implement \emph{CSDLDF} in CUDA.\footnote{Due to WebGPU’s 128MB storage buffer limit, our maximum test size is limited to $2^{25}$ 32-bit elements.}

Cross-API comparisons remain challenging, as despite being nearly identical at a high level, the CUDA implementation consistently outperforms its Dawn counterpart. Additionally, the performance cost of \emph{Decoupled Fallback} is most evident in our CUDA implementation when the entire input fits within the 2080 Super’s 4MB L2 cache, where increased synchronization and compute overhead slightly reduce throughput. However, at larger input sizes, where memory accesses extend beyond the cache, \emph{CSDLDF} achieves parity with CUB.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{graphics/cuda_plot.pdf}
  \caption{Performance versus CUB (NVIDIA 2080 Super)}
\end{figure}

@@

\begin{table*}
  \centering
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l r r r r r r}
    \toprule
    Metric & Intel HD620 & ARM Mali-G78 MP20 & Apple M3 & Apple M1 Max & Nvidia 2080 Super & AMD 7900 XT \\
    \midrule
    Avg \emph{Memcpy} (32-bit ele/sec $\cdot 10^9$)  & 1.50   & 2.48  & 10.25  & 38.62  & 54.39  & 72.28  \\
    Avg \emph{CSDLDF} (32-bit ele/sec $\cdot 10^9$)   & 1.49   & 2.79  & 10.22  & 38.84  & 53.29  & 94.60  \\
    Avg \emph{RTS} (32-bit ele/sec $\cdot 10^9$)      & 1.06   & 2.02  & 6.81   & 25.13  & 35.90  & 61.49  \\
    \emph{CSDLDF}/\emph{Memcpy}       & 99.1\%    & 112.8\%  & 99.8\%    & 100.6\%   & 98.0\%    & 130.9\%   \\
    \emph{CSDLDF}/\emph{RTS}          & 140.2\%   & 138.5\%  & 150.2\%   & 154.5\%   & 148.4\%   & 153.8\%   \\
    Avg Spins per Wg    & 0.305    & 1.477    & 0.606    & 0.604    & 0.419    & 0.148    \\
    Avg Lookback Length per Wg  & 1.726    & 3.537    & 3.415    & 10.782   & 13.046   & 20.722   \\
    Avg Total Fallbacks Initiated      & 254.662   & 2504.704  & 365.546   & 990.962   & 86.400   & 9.308   \\
    Avg Total Successful Insertions    & 6.170     & 122.436   & 4.612     & 0.012    & 0.284    & 0.020    \\
    \bottomrule
  \end{tabular*}
  \caption{\emph{Decoupled Fallback} performance, 32-bit Inclusive Prefix Sum, Input Size $\mathbf{2^{25}}$}
  \label{tab:results}
\end{table*}

As shown in Table~\ref{tab:results}, \emph{CSDLDF} achieves at least a $\sim$40\% speedup over \emph{RTS} across all tested devices—closely approaching the theoretically expected $O(3n)/O(2n)$ 50\% speedup, except on the ARM Mali. Since \emph{Memcpy} is purely memory-bandwidth-bound, \emph{CSDLDF} should ideally match its throughput. This holds across most devices, except for the AMD 7900 XT and ARM Mali, where \emph{CSDLDF} outperforms \emph{Memcpy}, suggesting that the additional compute latency counterintuitively improves memory access patterns by reducing paging. \john{I have no idea what this previous phrase means.} \thomas{In my head, if you read from location A, and then write immediately to B, you need room in your cache for A and B because the workgroup doesn't execute in lockstep. If you read, do just the right amount of work, then write B, you only need to hold A or B because the compute latency ensures that all the work from A is done before you get to B. Not sure if this is correct.}On the ARM Mali, \emph{CSDLDF} appears to underperform, achieving only a 36\% speedup instead of the expected 50\%. However, as its performance is nearly identical to \emph{Memcpy}, this indicates that the implementation is memory-bandwidth-bound, limiting further gains.

\subsection{Decoupled Fallback Statistics}
To better understand how differing devices interact with \emph{CSDLDF}, we also collect a number of statistics during its execution:
\begin{itemize}
  \item \textbf{Spins per Workgroup:} How many times does a workgroup make a waiting spin?
  \item \textbf{Lookback Length per Workgroup:} How many work tiles does a workgroup traverse to complete \emph{Chained-Propagation}?
  \item \textbf{Total Fallbacks Initiated:} What is the total number of fallback operations initiated across all workgroups?
  \item \textbf{Total Successful Insertions:} What is total number of fallback operations that successfully update a blocking (or late) tile to \emph{Ready}?
\end{itemize}
Despite significantly different absolute occupancies, the number of waiting spins remains consistent across devices. Likewise, lookback length is stable and remains well below the theoretical occupancy bound. Expressed as a percentage of occupancy, it remains under 10\% for all devices, except for the Intel HD620, where its elevated value is likely an artifact of its low absolute occupancy. Most notably, the devices lacking FPG---Apple M3 and M1 Max---do not exhibit significantly higher successful insertion rates than the others, suggesting that true deadlocking is a rare occurrence.

\subsection{Simulated Blocking}
To demonstrate the resilience of our technique to unfair schedulers, we simulate fallback rates significantly higher than those observed organically by intentionally forcing fallbacks through the omission of work tile reductions. Beginning with infrequent omissions---1 in 512 work tiles---we progressively increase the rate to near-continuous levels, reaching 1 in 2 work tiles.
\begin{figure*}[t]
  \centering
  \begin{subfigure}[t]{0.48\linewidth}
    \centering
    \includegraphics[width=\linewidth]{graphics/fallbacksInitiated_plot.pdf}
    \caption{Fallbacks Initiated}
    \label{fig:fallbacks_initiated}
  \end{subfigure}\hfill
  \begin{subfigure}[t]{0.48\linewidth}
    \centering
    \includegraphics[width=\linewidth]{graphics/totalSpins_plot.pdf}
    \caption{Total Spins}
    \label{fig:total_spins}
  \end{subfigure}

@@

As shown in Figure~\ref{fig:execution_time}, performance relative to a no-omission baseline remains above 85\% even at a 100\% fallback rate across all devices—except for the Intel HD620, where only a 25\% fallback rate is induced. Notably, this performance level remains well above the equivalent \emph{RTS} performance of 66\%. While Figures~\ref{fig:fallbacks_initiated} and \ref{fig:total_spins} confirm that we are inducing fallbacks, this test primarily measures the impact of increasing fallback operations rather than true blocking behavior. The finer details of this distinction are discussed next.
\section{Discussion}
\subsection{Limits on Speed-of-Light Performance}

In true blocking scenarios, the stalled workgroup is starved of progress, but execution resources are not lost; rather, they are redistributed to other workgroups. As a result, we would expect lookback lengths to be longer than those observed in our simulation.

@@

\subsection{Trade-offs}
Decoupled Fallback introduces several tradeoffs that must be considered: (1) Without a memory fence, the scan cannot be performed in place and requires a separate output buffer to prevent false-positive fallback operations on post-scanned values, as the tile state may not update before the scan values. (2) For monoids larger than 64 bits, multiple subgroups are required, increasing synchronization overhead due to the need for shared memory and barriers. (3) In certain in-situ compaction algorithms where fallback work-tiles cannot be easily reduced and must be held in registers, Decoupled Fallback may reduce occupancy. (4) Additionally, in compute-bound in-situ compaction algorithms, the cost of fallbacks may outweigh the savings from reduced global memory traffic. (5) Finally, because the control flow of \emph{Decoupled Fallback} introduces barriers latency incurred during lookback cannot be hidden by workgroup computation.

\subsection{Future Work}

\subsection{Conclusion}

@@

\clearpage
\appendix
\section{Artifact}
\subsection{Baseline Comparisons: \emph{Memcpy}}
\label{sec:memcpy}
\begin{table}[h!]
  \small
  \centering
  \begin{tabular}{l r r}
    \toprule
    Device                                & \makecell{Memory                   \\ Bandwidth \\ (GB/s)} & \makecell{Expected \\ Throughput \\ (32-bit ele/sec)} \\
    \midrule
    Intel HD620 (Single-/Dual-Channel)    & 17.1 / 34.1      & 2.13e9 / 4.26e9 \\
    ARM Mali-G78 MP20 / Pixel 6a (Tensor) & 51.2             & 6.40e9          \\
    Apple M3                              & 102.4            & 12.8e9          \\
    Apple M1 Max                          & 409.6            & 51.2e9          \\
    Nvidia 2080 Super                     & 496              & 62.0e9          \\
    AMD 7900 XT                           & 800              & 100.0e9         \\
    \bottomrule
  \end{tabular}
  \caption{Vendor-Specified Memory Bandwidth and Expected Throughput (32-bit Elements per Second)}
  \label{tab:memory_bandwidth}
\end{table}
Although all backend APIs support \emph{timestamp queries inside encoders}, Apple GPUs do not, so we are forced to implement our own \emph{Memcpy} kernel to measure memory bandwidth, as opposed to simply timing an API's copy buffer operation. It is well known that \emph{observed} peak memory bandwidth, tends to diverge from vendor-listed peak bandwidth, so for completeness we list these figures above. We make a number of observations:
\begin{itemize}
  \item \textbf{Intel HD620:} As our HD620 laptop is only equipped with a single memory DIMM, its peak bandwidth is limited to the lower bound, 2.13e9 ele/sec. Thus, our observed bandwidth of 1.498e9 is reasonable.
  \item \textbf{ARM Mali-G78 MP20 / Pixel 6a (Tensor):} The observed throughput is significantly lower than expected, despite a vendor-listed bandwidth of 6.40e9 ele/sec. The cause of this discrepancy remains unclear, but we note other developers have observed similar speeds.
  \item \textbf{Apple M1 Pro Max:} Observed throughput matches expectations.
  \item \textbf{Nvidia 2080 Super:} Observed throughput matches expectations.
  \item \textbf{AMD 7900 XT:} Our above result, where \emph{CSDLDF} was observed to be faster than \emph{Memcopy}, is likely due to compute latency counterintuitively improving memory access patterns, leading to reduced paging. Furthermore, we see that our \emph{CSDLDF} throughput is within spec and reasonable.
\end{itemize}
